---
html_document: default
author: "Lucia Patalacci"
date: "16 April 2019"
output:
  html_document:
    df_print: paged
pdf_document: default
title: 'Botanical Characterization and Geographical Origins in the Honey Market: a
  Bayesian Analysis based on Melissopalynological Features:
  Complete Analysis MPOL Data'
---
```{bash}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## DATA ANALYSIS

####Run Libraries
```{r library, include=FALSE}
library(tidyr)
library(plyr)
library(tibble)
library(caret)
library(ggplot2)
library(gmodels)
library(vcd)
library(dplyr)
library(rstan)
library(readxl)
library(psych)
library(cluster)
library(randomForest)
library(RColorBrewer)
```

##Data Loading
Load Excel files with table: MPOL analysis
```{r complete, include = FALSE}
#Data Loading #
#read xlsx data and save it into dataset
mpol_data<- read_excel("dataset_completo.xlsx", sheet="MPOL",col_names =TRUE,col_types=NULL,trim_ws=TRUE)
```

###Data Preparation
- Make data names legal
- Handle missing values: they are all set to zero value (absence of pollen type)
- Drop "ND" levels (entire row) from Geografical Macroarea.geo, and Area.geo
- Convert categorical features (Geografical Macroarea and Area, Botanical Origin)
- Remove non useful columns: Sample REF, Analysis TYPE and TOT_CONTA (100%)
- Round decimal values to 4 digits
- save new dataset in RData format

```{r CLEANDATA, include = FALSE}
#make data names legal##
names(mpol_data) <- make.names(names(mpol_data))

#Handle missing values##
mpol_data_new <- replace(mpol_data,is.na(mpol_data),0)

#Convert column from chr to factor
mpol_data_new [, c("Macroarea.geo", "Area.geo","Botanica.acc")]<- lapply(mpol_data_new [, c("Macroarea.geo", "Area.geo","Botanica.acc")],factor)

#check cleaning of datasets
#head(mpol_data_new,n=5)
#tail(mpol_data_new,n=5)

# Remove columns REF and TIPO, TOT_CONTA#
mpol_data_new$REF <- mpol_data_new$TIPO <- mpol_data_new$TOT_STIMA <-NULL
```
#####Grouping levels Macroarea.geo 
A new multi level feature has been created by grouping MacroArea.geo classes  to define classification models
```{r  RM_AREA_GROUP, echo=FALSE}
#Group geo
table(mpol_data_new$Macroarea.geo)
mpol_data_new <- data.frame(geo.group="geo.group", mpol_data_new, stringsAsFactors = FALSE)
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Miscela con Est Europa"] = "OtherEurope"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Est Europa"]= "OtherEurope"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Europa"] = "OtherEurope" 
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Italia"] = "Italy"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Africa"] = "OtherWorld"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Oceania"] = "OtherWorld"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Asia"] = "OtherWorld"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="Miscela"] = "OtherWorld"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="ND"] = "ND"
mpol_data_new$geo.group[mpol_data_new$Macroarea.geo=="America"] = "OtherWorld"
mpol_data_new$geo.group<- sapply(mpol_data_new [, "geo.group"],factor)
table(mpol_data_new$geo.group)
```

```{r  RM_AREA_GROUP2, echo=FALSE}
# elimino i livelli "ND"
mpol_data_new <-subset(mpol_data_new, subset = mpol_data_new$geo.group!="ND")
mpol_data_new$geo.group<-factor(mpol_data_new$geo.group)
table(mpol_data_new$geo.group)
```

```{r  CLEANDATA2, echo=FALSE}

# round decimals for numerical features
signif(mpol_data_new[,6:222],digits = 4)
```

#####Botanical.acc
A new multi level feature has been created by grouping Botanical.acc for higher frequencies classes to define classification models
```{r  RM_HONEY_TYPE, echo=FALSE}
#Levels in botanic.acc
table(mpol_data_new$Botanica.acc)
levels(mpol_data_new$Botanica.acc)

#############################################################
#''' ################my trials to group
#HoneyType<-as.data.frame(table(mpol_data_new$Botanica.acc))
#str(HoneyType)
#n<-nrow(HoneyType)
#{
#  HoneyType[1:n]
#for (i in n)
#  if (HoneyType[i]$Freq<=4)
#HoneyType$Var1[i]=="MieliRari"
#else HoneyType$Var1[i]
#}
#HoneyType<-as.data.frame(table(mpol_data_new$Botanica.acc))
#str(HoneyType)
#HoneyType <- data.frame(HoneyType="HoneyType", HoneyType, stringsAsFactors = FALSE)
#HoneyType[order(HoneyType$type),]
#HoneyType$HoneyType[HoneyType$type==1] = "RareHoney"
#HoneyType$HoneyType[HoneyType$type==2] = "RareHoney"
#HoneyType$HoneyType[HoneyType$type==3] = "RareHoney"
#HoneyType$HoneyType[HoneyType$type==4] = "RareHoney"
#'''
############################################################
```

```{r  RM_SAVE DATA, echo=FALSE}
#save datasets into a file RData#
save(mpol_data_new, file = "Dataset_MPOL.RData")
```

##Classification Models on MPOL Type Analysis
### Descriptive Analysis in MPOL dataset
```{r MPOL, include= FALSE}
# MPOL type Classification #
load("Dataset_MPOL.RData")

#####Data checking#######
#str(mpol_data_new)
#check TOT_STIMA column
#mpol_data_new$mpol_conta_tot <- rowSums(mpol_data_new[7:223])
#mpol_conta_tot <- rowSums(mpol_data_new[7:223])
#testdiff<- mpol_data_new$mpol_conta_tot - mpol_data_new$TOT_CONTA 
```

####Descriptive Statistical Analysis
```{r MPOL_DES, echo=FALSE}
#summary of descriptive statistics
describe(mpol_data_new)
```

####create new double-level variable group by geographical origin:italian vs foreign and botanical origin
A new double level feature has been created by grouping in "Italian" and "Foreign" classes the geographical assessed Macro area to define a classification model based on boolean feature

```{r MPOL_BOOL, echo= FALSE}
#geo.origin <- c()
mpol_data_new <- data.frame(geo.origin="geo.origin", mpol_data_new, stringsAsFactors = FALSE)
mpol_data_new$geo.origin[mpol_data_new$Macroarea.geo=="Italia"] = "Italian"
mpol_data_new$geo.origin[mpol_data_new$Macroarea.geo!="Italia"] = "Foreign"
mpol_data_new$geo.origin<- sapply(mpol_data_new [, "geo.origin"],factor)
table(mpol_data_new$geo.origin)

```

```{r ORDERED_GEO_ORIGIN, include = FALSE}
# re-order levels
reorder_size <- function(x) {
  factor(x, levels = names(sort(table(x), decreasing = TRUE)))
}
```

###Botanical species frequencies by geo origin: Italian and Foreign
Below, the frequencies of single botanical types distributed by "Italian" and "Foreign" declared origin and the corresponding mosaic chart that can allow to visualize the data from contingency table: each rectangle and its height represent the proportional value of Honey type by countries.
```{r MOSAICPLOT, echo = FALSE}
geo_botanic <-xtabs(~ reorder_size(Botanica.acc) +geo.origin, mpol_data_new)
geo_botanic
geo_botanic1 <-xtabs(~  geo.origin + reorder_size(Botanica.acc), mpol_data_new)

#par(mar=c(5, 4, 4, 5) + 0.1)    # Widen right margin slightly, to fit horizontal labels
mosaicplot(geo_botanic1, main = "Honey type with higher frequency by Country",
           xlab = "Countries",
           ylab = "Botanical Type",
           las = 1,
           col=brewer.pal(n=length(mpol_data_new$Botanica.acc),name = "Paired"))
```
####Some histograms:
```{r PLOTBOTANICFREQ, echo=FALSE}
# Histogram on a Categorical variable
geo1 <- ggplot(mpol_data_new,aes(x=geo.origin))
geo1 + geom_bar(aes(fill=(reorder_size(mpol_data_new$Honey.type))), width = 0.5) + 
  theme(axis.text.x = element_text(angle=0, vjust=0.6),
        axis.text.y = element_text(angle=90, vjust=0.6),legend.position="None") + 
  labs(title="Honey Type Density",
         subtitle="Honey Types Grouped by Italian and Foreign Origin",
         caption="Source: Mpol - Complete Analysis",
         x="Countries",
         fill="Botanical Origin")
```
###Botanical species and Geographical frequencies histograms
Below the graphical representation of Geographical distribution histogram, Botanical distribution histogram, respectively ordered by frequencies 
```{r AREAPLOT, echo = FALSE}
# Histogram on a Categorical variable
geo2 <- ggplot(mpol_data_new,aes(x=reorder_size(Macroarea.geo)))
geo2 + geom_bar(aes(fill=(reorder_size(mpol_data_new$Honey.type))), width = 0.5) + 
  theme(axis.text.x = element_text(angle=0, vjust=0.6),
        axis.text.y = element_text(angle=90, vjust=0.6),legend.position="None") + 
  labs(title="Honey Type Density",
         subtitle="Honey Types Grouped by Geographical MacroArea",
         caption="Source: Mpol - Complete Analysis",
         x="MacroArea",
         fill="Botanical Origin")
```

```{r BOTANICAL_ORIGIN, echo = FALSE}
#barplot by Botanical Origin #####
# Histogram on a Categorical variable
bot1 <- ggplot(mpol_data_new,aes(x=reorder_size(Botanica.acc)))
bot1 + geom_bar(aes(fill=reorder_size(Macroarea.geo)), width = 0.5) + 
  theme(axis.text.x = element_text(angle=90, vjust=0.6),
        axis.text.y = element_text(angle=90, vjust=0.6),
        legend.title = element_text(size=12, color = "salmon", face="bold"),
        legend.justification=c(0,1), 
        legend.position=c(0.80, 0.95),
        legend.background = element_blank(),
        legend.key = element_blank()) + 
  labs(title="Honey Type Distribution",
         legend.position="Top-Left",
         subtitle="Honey Types Grouped by Geographical MacroArea",
         caption="Source: Mpol - Complete Analysis",
         x="Botanical Origin",
         fill="MacroArea")
```
Below, the scatterplot chart of botanical types distributed respectively by double classes and by geographical macro-area
```{r GEORIGIN.VS.HONEY, echo=FALSE}
geo_bot<-ggplot(mpol_data_new, aes(x=Botanica.acc, y=Macroarea.geo)) + geom_point()+
  geom_point(aes(col=Botanica.acc), size=1)+
  theme(axis.text.x = element_text(angle = 90, hjust =1, size=7),legend.position="None")+
  labs(title="Honey Type by MacroArea", y="", x="Botanical Origin", caption="Source: Mpol - Complete Analysis")
plot(geo_bot)
```
##Random Forest Models for classification

###Random Forest with class "Italian" and "Foreign" on boolean pollen features (pollen presence/not)
####Data Transformation and splitting data in training, validation and test sets
```{r  RANDOMFOREST_BOOL, echo = FALSE }
#copy dataframe mpol_data_new#
mpol_data_bool<-mpol_data_new
#convert frequencies in boolean (0,1)#
mpol_data_bool[,6:223][mpol_data_bool[,6:223]>0.001] <- 1
mpol_data_bool[,6:223][mpol_data_bool[,6:223]<=0.001] <- 0

#training, validation and test set#
set.seed(2205)
mpol_data_split.b <- sample(1:2, size=nrow(mpol_data_bool), prob=c(0.70,0.30), replace = TRUE)
mpol_train.b <- mpol_data_bool[mpol_data_split.b==1,]
mpol_test.b <- mpol_data_bool[mpol_data_split.b==2,]
```
####Random Forest Model and Confusion Matrix m=sqrt(218)

```{r RMBOOL_GEO_15, echo=FALSE}
set.seed(123)
mpol_rf_classifier.b15 = randomForest(geo.origin ~.-Macroarea.geo -Area.geo -geo.group -Honey.type, data = mpol_train.b, ntree=500, mtry=sqrt(218), importance=TRUE, proximity=TRUE)
mpol_rf_classifier.b15
```
####Random Forest Model and Confusion Matrix m=(218/2)
```{r  RMBOOL_GEO_108, echo=FALSE}
#exclude categorical variables (RF model do not accept more than 53)#
set.seed(5674)
mpol_rf_classifier.b108 = randomForest(geo.origin ~.-Macroarea.geo -Area.geo -geo.group -Honey.type, data = mpol_train.b, ntree=500, mtry=(218/2), importance=TRUE, proximity=TRUE)
mpol_rf_classifier.b108
```

####Random Forest Model and Confusion Matrix m=3
```{r  RMBOOL_GEO_03, echo=FALSE}
#exclude categorical variables (RF model do not accept more than 53)#
set.seed(2345)
mpol_rf_classifier.b3 = randomForest(geo.origin ~.-Macroarea.geo -Area.geo -geo.group -Honey.type, data = mpol_train.b, ntree=500, mtry=3, importance=TRUE, proximity=TRUE)
mpol_rf_classifier.b3
```

```{r  RMBOOL_GEO_PLOT, echo=FALSE}
layout(matrix(c(1,2),nrow=1),
       width=c(4,1)) 
par(mar=c(5,4,4,0)) #No margin on the right side
plot(mpol_rf_classifier.b15, log="y", main="RF Classifier Boolean with 15 variables")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("topleft", colnames(mpol_rf_classifier.b15$err.rate),col=1:4,cex=0.8,fill=1:4)
```

######Pollen Type importance in classification
```{r  RMBOOL_GEO15 IMPORTANCE, echo=FALSE}
#importance(mpol_rf_classifier.b15)
varImpPlot(mpol_rf_classifier.b15,main="Pollen Type Importance for Class:Italian vs Foreign - 15 var") 
```

######Pollen Type importance in classification
```{r  RMBOOL_GEO108 IMPORTANCE, echo=FALSE}
#importance(mpol_rf_classifier.b108)
varImpPlot(mpol_rf_classifier.b108,main="Pollen Type Importance for Class:Italian vs Foreign 108 var")
```

######Pollen Type importance in classification
```{r  RMBOOL_GEO3 IMPORTANCE, echo=FALSE}
#importance(mpol_rf_classifier.b3)
varImpPlot(mpol_rf_classifier.b3,main="Pollen Type Importance for Class:Italian vs Foreign 3 var")
```

### Validation set assessment on Random Forest on Classes "Italian" and "Foreign": looking at confusion matrix
##### RF 15 predictors
```{r  RMBOOL_VALID15, echo=FALSE}
validation_RF_hat.b <- predict(mpol_rf_classifier.b15, newdata=mpol_test.b[,-1], type="class")
mean(validation_RF_hat.b == mpol_test.b[,1])
confusionMatrix(mpol_test.b[,1],validation_RF_hat.b)
```
##### RF 109 predictors
```{r  RMBOOL_VALID108, echo=FALSE}
validation_RF_hat.b1 <- predict(mpol_rf_classifier.b108, newdata=mpol_test.b[,-1], type="class")
mean(validation_RF_hat.b1 == mpol_test.b[,1])
confusionMatrix(mpol_test.b[,1],validation_RF_hat.b1)
```
##### RF 3 predictors
```{r  RMBOOL_VALID03, echo=FALSE}
validation_RF_hat.b2 <- predict(mpol_rf_classifier.b3, newdata=mpol_test.b[,-1], type="class")
mean(validation_RF_hat.b == mpol_test.b[,1])
confusionMatrix(mpol_test.b[,1],validation_RF_hat.b2)
```

###Random Forest on Geografical Classification (Feature"geo.group") 
##### RF 15 predictors geoarea
```{r RF_GROUP15, echo=FALSE}
##exclude categorical variables (RF model do not accept more than 53)#
set.seed(3489)
mpol_rf_classifier.b15g = randomForest(geo.group ~.-Macroarea.geo -Area.geo -geo.origin -Honey.type, data=mpol_train.b, ntree=500, mtry=sqrt(218), importance=TRUE,proximity=TRUE)
mpol_rf_classifier.b15g
```
##### RF 109 predictors geoarea
```{r  RF108_GROUP, echo=FALSE}
##exclude categorical variables (RF model do not accept more than 53)#
set.seed(2001)
mpol_rf_classifier.b108g = randomForest(geo.group ~.-Macroarea.geo -Area.geo -geo.origin -Honey.type, data =  mpol_train.b, ntree=500, mtry=218/2, importance=TRUE,proximity=TRUE)
mpol_rf_classifier.b108g
```
##### RF 3 predictors geoarea
```{r  RF03_GROUP, echo=FALSE}
##exclude categorical variables (RF model do not accept more than 53)#
set.seed(4553)
mpol_rf_classifier.b3g = randomForest(geo.group ~.-Macroarea.geo -Area.geo -geo.origin -Honey.type, data = mpol_train.b, ntree=500, mtry=3, importance=TRUE,proximity=TRUE)
mpol_rf_classifier.b3g
```

```{r  RANDOMFORESTMOD_GROUP_PLOT, echo=FALSE}
layout(matrix(c(1,2),nrow=1),
       width=c(4,1)) 
par(mar=c(5,4,4,0)) #No margin on the right side
plot(mpol_rf_classifier.b15g, log="y", main="RF Classifier Class:Geo Macro Area with 15 var" )
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("topleft", colnames(mpol_rf_classifier.b15g$err.rate),col=1:11,cex=0.8,fill=1:4)
```

```{r  RANDOMFORESTMOD_AREA_IMPORTANCE, echo=FALSE}
#importance(mpol_rf_classifier.b1)
varImpPlot(mpol_rf_classifier.b15g, main="Pollen Type Importance Class:Geo Macro Area with 15 vars")
```

### Validation set assessment on Random Forest on Classes "Italian" and "Foreign": looking at confusion matrix
##### RF 15 predictors
```{r  RMBOOLAREA_VALID15, echo=FALSE}
validation_RF_hat.bg <- predict(mpol_rf_classifier.b15g, newdata=mpol_test.b[,-2], type="class")
mean(validation_RF_hat.bg == mpol_test.b[,2])
confusionMatrix(mpol_test.b[,2],validation_RF_hat.bg)
```
##### RF 109 predictors
```{r  RMBOOLAREA_VALID108, echo=FALSE}
validation_RF_hat.b1g <- predict(mpol_rf_classifier.b108g, newdata=mpol_test.b[,-2], type="class")
mean(validation_RF_hat.b1g == mpol_test.b[,2])
confusionMatrix(mpol_test.b[,2],validation_RF_hat.b1g)
```
##### RF 3 predictors
```{r  RMBOOLAREA_VALID03, echo=FALSE}
validation_RF_hat.b2g <- predict(mpol_rf_classifier.b3g, newdata=mpol_test.b[,-2], type="class")
mean(validation_RF_hat.b2g == mpol_test.b[,2])
confusionMatrix(mpol_test.b[,2],validation_RF_hat.b2g)
```
###Random Forest with typeuencies
####Random Forest on Classes "Italian" and "Foreign"
#####Splitting data in training, validation and test sets
```{r  RMtype_GEO15_SPLIT, echo=FALSE}
#training, validation and test set#
set.seed(2205)
mpol_data_split <- sample(1:2, size=nrow(mpol_data_new), prob=c(0.7,0.3), replace = TRUE)
mpol_train <- mpol_data_new[mpol_data_split==1,]
mpol_test <- mpol_data_new[mpol_data_split==2,]
```
#####RF 15 PREDICTORS
```{r  RMFREQ_BOOL15, echo=FALSE}
#exclude categorical variables (RF model do not accept more than 53)#
set.seed(123)
mpol_rf_classifier.f15 = randomForest(geo.origin ~ .-Macroarea.geo -geo.group -Area.geo -Honey.type, data = mpol_train, ntree=500, mtry=sqrt(218), importance=TRUE,proximity=TRUE)
mpol_rf_classifier.f15
```
#####RF 109 PREDICTORS
```{r  RMFREQ_GEO108, echo=FALSE}
#exclude categorical variables (RF model do not accept more than 53)#
set.seed(123)
mpol_rf_classifier.f108 = randomForest(geo.origin ~ .-Macroarea.geo -geo.group -Area.geo -Honey.type, data = mpol_train, ntree=500, mtry=2180/2, importance=TRUE,proximity=TRUE)
mpol_rf_classifier.f108
```
#####RF 3 PREDICTORS
```{r  RMFREQ_GEO3, echo=FALSE}
#exclude categorical variables (RF model do not accept more than 53)#
set.seed(123)
mpol_rf_classifier.f3 = randomForest(geo.origin ~ .-Macroarea.geo -geo.group -Area.geo -Honey.type, data = mpol_train, ntree=500, mtry=3, importance=TRUE,proximity=TRUE)
mpol_rf_classifier.f3
```

```{r  RMFREQ_GEO15, echo=FALSE}
layout(matrix(c(1,2),nrow=1),
       width=c(4,1)) 
par(mar=c(5,4,4,0)) #No margin on the right side
plot(mpol_rf_classifier.f15, log="y", main="RF Classifier on Frequency class:Italian vs Foreign 15 var ")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("topleft", colnames(mpol_rf_classifier.f15$err.rate),col=1:4,cex=0.8,fill=1:4)
```

#### Features importance on RF Model with classes Italian and Foreign
```{r  RMFREQ_GEO15_IMPORTANCE, echo=FALSE}
#importance(mpol_rf_classifier)
varImpPlot(mpol_rf_classifier.f15, main="Pollen Type Frequency Importance on Classes 15 var")
```
### Validation set assessment on Random Forest on Classes "Italian" and "Foreign": looking at confusion matrix
```{r  RMFREQ_GEO15_VALID, echo=FALSE}
validation_RF_hat.f <- predict(mpol_rf_classifier.f15, newdata=mpol_test[,-1], type="class")
mean(validation_RF_hat.f == mpol_test[,1])
confusionMatrix(mpol_test[,1],validation_RF_hat.f)
```

```{r  RMFREQ_GEO108_VALID, echo=FALSE}
validation_RF_hat.f1 <- predict(mpol_rf_classifier.f108, newdata=mpol_test[,-1], type="class")
mean(validation_RF_hat.f1 == mpol_test[,1])
confusionMatrix(mpol_test[,1],validation_RF_hat.f1)
```

```{r  RMFREQ_GEO3_VALID, echo=FALSE}
validation_RF_hat.f2 <- predict(mpol_rf_classifier.f3, newdata=mpol_test[,-1], type="class")
mean(validation_RF_hat.f2 == mpol_test[,1])
confusionMatrix(mpol_test[,1],validation_RF_hat.f2)
```

####RANDOM FOREST WITH 10-FOLD CV 10 TIMES
```{r  RANDOMFORESTMOD_10CV, echo=FALSE}
ctrl <-trainControl(method="repeatedcv",
                    number = 10, repeats = 10)
grid_rf <- expand.grid(.mtry = c(2,5,10,15,20))

set.seed(2402)
cv_rf_classifier_10<- train(geo.origin ~ .-Macroarea.geo -geo.group -Area.geo -Honey.freq, data = mpol_data_new, method = "rf",
                         metric = "Kappa", trControl = ctrl, 
                         tuneGrid = grid_rf)
cv_rf_classifier_10
```
#####Plot of RF classifier 10_CV 10 times 
```{r  RANDOMFORESTPLOT_10CV, echo=FALSE}
#layout(matrix(c(1,2),nrow=1),
       #width=c(4,1)) 
#par(mar=c(5,4,4,0)) #No margin on the right side
plot(cv_rf_classifier_10, log="y", main="Kappa Measure vs No.of Selected Predictors")
#par(mar=c(5,0,4,2)) #No margin on the left side
#plot(c(0,1),type="n", axes=F, xlab="", ylab="")
```